New and simplified manual controls for projection and slice tours, with application to exploring classification boundaries in high dimensions

### 2022-10-04 paper/appendix.Rmd
Spellcheck and typos

- `data`: A data matrix, potentially with grouping that has a label (string) and a flag  (numerical) in the last two columns.
- `projmat`: The initial projection matrix. It is possible to start with a random projection with the input "random" (including the quotes) in this entry field.
- `height`: The initial height of the slice.
- `heightRange`: The range of slice heights to be explored.
- `legendQ`: A flag signaling whether the data matrix includes a column with group labels (1) or not (0).
- `flagQ`: A flag signaling whether the data matrix includes a numerical flag for the groups.
- `colorFunc`: specifies the colour mapping for the different groups.

@@

The function renders a manual tour with controls that allow the user to navigate through different projections. For a given projection, a slider permits the user to change the slice width within its range and a separate box switches the view to a projection. The user can also change the centre point of slices; the size of the points rendered and the scale of the plot through sliders. An additional box displays the coordinates of the projection matrix explicitly.

This function can be used for ungrouped data by setting legendQ and flagQ to 0.
To remove a group from the visual display, the user can choose specific colours for the entry colorFunc. For example, {Red,White,Green} would display groups 1 (red) and 3 (green) while making 2 invisible.

@@

- `data`: A data matrix with one or more groups labeled by a flag (numerical, last column).
- `projmat`: The initial projection matrix. It is possible to start with a random projection with the input "random" (including the quotes) in this entry field.
- `legendNames`: an optional list of labels for the groups.
- `colorFunc`: specifies the colour mapping for the different groups.

`ProjectedLocatorPlot` displays the interactive controls slightly differently. Several locator panes are displayed above the display of the projected data and each one corresponds to a row in the projection matrix. The updating behaviour of the projection is the same as in `ProjectionPlot`. The function and its arguments are given below:

@@

- `data`: A data matrix with one or more groups labeled by a flag  (numerical, last column). 
- `projmat`: The initial projection matrix. It is possible to start with a random projection with the input "random" (including the quotes) in this entry field.
- `legendNames`: an optional list of labels for the groups.
- `colorFunc`: specifies the colour mapping for the different groups.

Finally `VisualiseSliceDynamic` allows to discern which points exist inside the slice and outside it. This function differs from `SliceDynamic` in that it looks at only one data set but displays both the points inside and outside the slice. The display panel allows the user to specify the size of both sets of points (inside and outside) along with the slice height, centre point and zoom level. The function and its arguments are given below:

>>>>

- `data`: A data matrix, potentially with grouping that has a label (string) and a flag  (numerical) in the last two columns.
- `projmat`: The initial projection matrix. It is possible to start with a random projection with the input "random" (including the quotes) in this entry field.
- `height`: The initial height of the slice.
- `heightRange`: The range of slice heights to be explored.
- `legendQ`: A flag signaling whether the data matrix includes a column with group labels (1) or not (0).
- `flagQ`: A flag signaling whether the data matrix includes a numerical flag for the groups.
- `colorFunc`: specifies the color mapping for the different groups.

@@

The function renders a manual tour with controls that allow the user to navigate through different projections. For a given projection, a slider permits the user to change the slice width within its range and a separate box switches the view to a projection. The user can also change the center point of slices; the size of the points rendered and the scale of the plot through sliders. An additional box displays the coordinates of the projection matrix explicitly.

This function can be used for ungrouped data by setting legendQ and flagQ to 0.
To remove a group from the visual display, the user can choose specific colors for the entry colorFunc. For example, {Red,White,Green} would display groups 1 (red) and 3 (green) while making 2 invisible.

@@

- `data`: A data matrix with one or more groups labeled by a flag (numerical, last column).
- `projmat`: The initial projection matrix. It is possible to start with a random projection with the input "random" (including the quotes) in this entry field.
- `legendNames`: an optional list of labels for the groups.
- `colorFunc`: specifies the color mapping for the different groups.

`ProjectedLocatorPlot` displays the interactive controls slightly differently. Several locator panes are displayed above the display of the projected data and each one corresponds to a row in the projection matrix. The updating behavior of the projection is the same as in `ProjectionPlot`. The function and its arguments are given below:

@@

- `data`: A data matrix with one or more groups labeled by a flag  (numerical, last column). 
- `projmat`: The initial projection matrix. It is possible to start with a random projection with the input "random" (including the quotes) in this entry field.
- `legendNames`: an optional list of labels for the groups.
- `colorFunc`: specifies the color mapping for the different groups.

Finally `VisualiseSliceDynamic` allows to discern which points exist inside the slice and outside it. This function differs from `SliceDynamic` in that it looks at only one data set but displays both the points inside and outside the slice. The display panel allows the user to specify the size of both sets of points (inside and outside) along with the slice height, center point and zoom level. The function and its arguments are given below:

### 2022-10-04 paper/appendix.tex
Spellcheck and typos

\begin{itemize}
\tightlist
\item
  \texttt{data}: A data matrix, potentially with grouping that has a
  label (string) and a flag (numerical) in the last two columns.
\item
  \texttt{projmat}: The initial projection matrix. It is possible to
  start with a random projection with the input ``random'' (including
  the quotes) in this entry field.
\item
  \texttt{height}: The initial height of the slice.
\item
  \texttt{heightRange}: The range of slice heights to be explored.
\item
  \texttt{legendQ}: A flag signaling whether the data matrix includes a
  column with group labels (1) or not (0).
\item
  \texttt{flagQ}: A flag signaling whether the data matrix includes a
  numerical flag for the groups.
\item
  \texttt{colorFunc}: specifies the colour mapping for the different
  groups.
\end{itemize}

@@

The function renders a manual tour with controls that allow the user to
navigate through different projections. For a given projection, a slider
permits the user to change the slice width within its range and a
separate box switches the view to a projection. The user can also change
the centre point of slices; the size of the points rendered and the
scale of the plot through sliders. An additional box displays the
coordinates of the projection matrix explicitly.

This function can be used for ungrouped data by setting legendQ and
flagQ to 0. To remove a group from the visual display, the user can
choose specific colours for the entry colorFunc. For example,
\{Red,White,Green\} would display groups 1 (red) and 3 (green) while
making 2 invisible.

@@

\begin{itemize}
\tightlist
\item
  \texttt{data}: A data matrix with one or more groups labeled by a flag
  (numerical, last column).
\item
  \texttt{projmat}: The initial projection matrix. It is possible to
  start with a random projection with the input ``random'' (including
  the quotes) in this entry field.
\item
  \texttt{legendNames}: an optional list of labels for the groups.
\item
  \texttt{colorFunc}: specifies the colour mapping for the different
  groups.
\end{itemize}

\texttt{ProjectedLocatorPlot} displays the interactive controls slightly
differently. Several locator panes are displayed above the display of
the projected data and each one corresponds to a row in the projection
matrix. The updating behaviour of the projection is the same as in
\texttt{ProjectionPlot}. The function and its arguments are given below:

@@

\begin{itemize}
\tightlist
\item
  \texttt{data}: A data matrix with one or more groups labeled by a flag
  (numerical, last column).
\item
  \texttt{projmat}: The initial projection matrix. It is possible to
  start with a random projection with the input ``random'' (including
  the quotes) in this entry field.
\item
  \texttt{legendNames}: an optional list of labels for the groups.
\item
  \texttt{colorFunc}: specifies the colour mapping for the different
  groups.
\end{itemize}

Finally \texttt{VisualiseSliceDynamic} allows to discern which points
exist inside the slice and outside it. This function differs from
\texttt{SliceDynamic} in that it looks at only one data set but displays
both the points inside and outside the slice. The display panel allows
the user to specify the size of both sets of points (inside and outside)
along with the slice height, centre point and zoom level. The function
and its arguments are given below:

>>>>

\begin{itemize}
\tightlist
\item
  \texttt{data}: A data matrix, potentially with grouping that has a
  label (string) and a flag (numerical) in the last two columns.
\item
  \texttt{projmat}: The initial projection matrix. It is possible to
  start with a random projection with the input ``random'' (including
  the quotes) in this entry field.
\item
  \texttt{height}: The initial height of the slice.
\item
  \texttt{heightRange}: The range of slice heights to be explored.
\item
  \texttt{legendQ}: A flag signaling whether the data matrix includes a
  column with group labels (1) or not (0).
\item
  \texttt{flagQ}: A flag signaling whether the data matrix includes a
  numerical flag for the groups.
\item
  \texttt{colorFunc}: specifies the color mapping for the different
  groups.
\end{itemize}

@@

The function renders a manual tour with controls that allow the user to
navigate through different projections. For a given projection, a slider
permits the user to change the slice width within its range and a
separate box switches the view to a projection. The user can also change
the center point of slices; the size of the points rendered and the
scale of the plot through sliders. An additional box displays the
coordinates of the projection matrix explicitly.

This function can be used for ungrouped data by setting legendQ and
flagQ to 0. To remove a group from the visual display, the user can
choose specific colors for the entry colorFunc. For example,
\{Red,White,Green\} would display groups 1 (red) and 3 (green) while
making 2 invisible.

@@

\begin{itemize}
\tightlist
\item
  \texttt{data}: A data matrix with one or more groups labeled by a flag
  (numerical, last column).
\item
  \texttt{projmat}: The initial projection matrix. It is possible to
  start with a random projection with the input ``random'' (including
  the quotes) in this entry field.
\item
  \texttt{legendNames}: an optional list of labels for the groups.
\item
  \texttt{colorFunc}: specifies the color mapping for the different
  groups.
\end{itemize}

\texttt{ProjectedLocatorPlot} displays the interactive controls slightly
differently. Several locator panes are displayed above the display of
the projected data and each one corresponds to a row in the projection
matrix. The updating behavior of the projection is the same as in
\texttt{ProjectionPlot}. The function and its arguments are given below:

@@

\begin{itemize}
\tightlist
\item
  \texttt{data}: A data matrix with one or more groups labeled by a flag
  (numerical, last column).
\item
  \texttt{projmat}: The initial projection matrix. It is possible to
  start with a random projection with the input ``random'' (including
  the quotes) in this entry field.
\item
  \texttt{legendNames}: an optional list of labels for the groups.
\item
  \texttt{colorFunc}: specifies the color mapping for the different
  groups.
\end{itemize}

Finally \texttt{VisualiseSliceDynamic} allows to discern which points
exist inside the slice and outside it. This function differs from
\texttt{SliceDynamic} in that it looks at only one data set but displays
both the points inside and outside the slice. The display panel allows
the user to specify the size of both sets of points (inside and outside)
along with the slice height, center point and zoom level. The function
and its arguments are given below:

### 2022-10-04 paper/paper.Rmd
Spellcheck and typos

---
title: |
  New and simplified manual controls for projection and slice tours, with application to exploring classification boundaries in high dimensions
type: Short Technical Note
author:
  - name: Ursula Laa
    affil: a
    email: ursula.laa@boku.ac.at
  - name: Alex Aumann
    affil: b
    email: aaum0002@student.monash.edu
  - name: Dianne Cook
    affil: c
    email: dicook@monash.edu
  - name: German Valencia
    affil: b
    email: german.valencia@monash.edu
affiliation:
  - num: a
    address: |
      Institute of Statistics, University of Natural Resources and Life Sciences, Vienna
  - num: b
    address: |
      School of Physics and Astronomy, Monash University
  - num: c
    address: |
      Department of Econometrics and Business Statistics, Monash University
bibliography: biblio.bib
geometry: margin=2.5cm
abstract: | 
  This paper describes new user controls for examining high-dimensional data using low-dimensional linear projections and slices. A user can interactively change the contribution of a given variable to a low-dimensional projection, which is useful for exploring the sensitivity of structure to particular variables. The user can also interactively shift the center of a slice, for example, to explore how structure changes in local subspaces. Two Mathematica notebooks are provided, which contains functions enabling the user to experiment with these new manual controls, with one specifically for exploring regions and boundaries produced by classification models. The advantage of Mathematica is its linear algebra capabilities, and interactive cursor location controls. Some limited implementation has also been made available in the R package \tt{tourr}. 
keywords: |
  data visualisation; grand tour; statistical computing; statistical graphics; multivariate data; dynamic graphics
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
  \usepackage{graphicx}
  \usepackage{nicematrix}
  \NiceMatrixOptions{code-for-first-row = \color{red} ,code-for-last-row = \color{red} ,code-for-first-col = \color{blue} ,code-for-last-col = \color{blue}}
output: rticles::tf_article
keep_tex: true
---

@@

<!-- Background -->
From a statistical perspective 3D is a rare data dimension, so unlike in most 3D rotation computer graphics applications, the more useful methods for data analysis need to work for arbitrary dimension. A good approch is to show projections from an arbitrary dimensional space to create dynamic data visualizations called *tours*. Tours involve views of high-dimensional ($p$) data with low-dimensional ($d$) projections. In his original paper on the grand tour, @As85 provided several algorithms for tour paths that could theoretically show the viewer the data *from all sides*. Prior to Asimov's work, there were numerous preparatory developments including @tukey's PRIM-9. PRIM-9 had user-controlled rotations on coordinate axes, allowing one to manually tour through low-dimensional projections. (A video illustrating the capabilities is available through video library of @ASA22.) Steering through all possible projections is impossible, unlike Asimov's tours which allows one to quickly see many, many different projections. After Asimov there have been many tour developments, which are summarized in @lee2021.

@@

Another potentially useful manual control, is to allow the user to choose the position of the center of a slice. The slice tour was introduced in @slicetour. It operates by converting the projection plane into a slice, by removing or de-emphasizing points that are further than a fixed orthogonal distance from the plane. The projection plane is usually thought of as passing through the center of the data. Manual control would allow the user to change the value of center point, by shifting it along a coordinate axis, while keeping the orientation of the projection plane fixed. The purpose would be to explore how or if the shape of the data, in the space orthogonal to the projection, changes as one gets away from the center. It would also allow the user to interactively decide on the thickness of the slice.

@@

The methods for updating component values in @cook_manual_1997 (and utilised in @spyrison_spinifex_2020) are prescribed primarily for a 2D projection, to take advantage of (then) newly developed 3D trackball controls made available for computer gaming. The first step was to construct a 3D manipulation space from a 2D projection. In this space, the coefficient of the controlled variable ranges between -1 and 1. Movements of a cursor are recorded and converted into changes in the values of $V_m$ thus changing the displayed 2D projection. The algorithm also provided constraints to horizontal, vertical, radial or angular motions only. The construction of the manipulation space overly complicates the manual controls, especially when considering possible techniques that will apply to arbitrary $d$.

@@

The new approach emerged from experiments on the tour using the linear algebra capabilities, and relatively new interactive graphics interface, available in @Mathematica. The components corresponding to $V_m$ are directly controlled by cursor movement, which updates row $m$ of $A$. The updated matrix is then orthonormalised.

@@

Mathematica provides much utility and versatility, such as inbuilt data visualisation, data manipulation and analysis, dynamic functionality, and symbolic and numeric computation. Most of this inbuilt functionality is user-friendly and described in the Mathematica documentation; typically, numerous examples are provided within the documentation, and some possible issues are outlined there.

@@

The examples presented in the application section below (and attached in notebook format in the supplementary material) all use our primary new function, `SliceDynamic`. This function typically accepts grouped data in the form of a matrix where the second last column details the name of the group and the last column details the group index (which can be one if there is only one group). After specifying the initial slice thickness and the slice range, the user is presented with an interactive display in which the control objects appear on the left and the slice visualisation appears on the right. The user can change the orientation of the slice via the locator pane, which changes the projection matrix;  a slider controls the slice thickness; and there is an input field, which changes the slice centre. The user can also change the appearance of the plot by zooming into the center or changing the point size with the sliders provided. It is worth noting that this zoom works best with data that has been centred and scaled. The projected data can be displayed to contrast it with the slice by ticking a box. The explicit projection matrix can also be displayed via another checkbox. This is especially useful when we wish to import a projection that was identified as interesting in the manual exploration into a later study, as shown in the applications below.

@@

    - `data`: A data matrix
    - `projmat`: The inital projection matrix
    - `centrepoint`: The initial position of the slice
    - `height`: The initial height of the slice
    - `heightRange`: The range of heights
    - `ptSize1`: The size of the points that exist within the slice
    - `ptSize2`: The size of the points that exist outside of the slice

@@

This function uses a different display for the projection that I find very useful. However as it stands it is specific to one data set. Needs to be generalised to arbitrary data if we want to keep it.

@@

We start by exploring the projections of the model prediction. Figure \ref{proj1} summarises the process. Rotating manually the view, we can visualise the location of predictions for each of the three species, and also get a sense of the difference between the two models. 
To illustrate this difference, we have manually rotated the projection for the RF model (left plot) to identify a view that shows the non-linear but block-type structure that is typical of this type of model. This particular projection ($A_1$) is exported so that it can later be used to show the LDA model (middle plot) and the actual data (right plot). What can be seen is the linearity of the LDA model, where the boundaries are linear and oblique to the variable axes. And, interestingly, this particular projection of the original data shows very distinct clusters of the three species. That means, the obscuring of the boundaries between groups for both of the models is driven by what is happening in the orthogonal space to the plane of the selected projection.

>>>>

---
title: |
  New and simplified manual controls for projection and slice tours, with application to exploring classification boundaries in high dimensions
type: Short Technical Note
author:
  - name: Ursula Laa
    affil: a
    email: ursula.laa@boku.ac.at
  - name: Alex Aumann
    affil: b
    email: aaum0002@student.monash.edu
  - name: Dianne Cook
    affil: c
    email: dicook@monash.edu
  - name: German Valencia
    affil: b
    email: german.valencia@monash.edu
affiliation:
  - num: a
    address: |
      Institute of Statistics, University of Natural Resources and Life Sciences, Vienna
  - num: b
    address: |
      School of Physics and Astronomy, Monash University
  - num: c
    address: |
      Department of Econometrics and Business Statistics, Monash University
bibliography: biblio.bib
geometry: margin=2.5cm
abstract: | 
  This paper describes new user controls for examining high-dimensional data using low-dimensional linear projections and slices. A user can interactively change the contribution of a given variable to a low-dimensional projection, which is useful for exploring the sensitivity of structure to particular variables. The user can also interactively shift the center of a slice, for example, to explore how structure changes in local subspaces. Two Mathematica notebooks are provided, which contain functions enabling the user to experiment with these new manual controls, with one specifically for exploring regions and boundaries produced by classification models. The advantage of Mathematica is its linear algebra capabilities, and interactive cursor location controls. Some limited implementation has also been made available in the R package \tt{tourr}. 
keywords: |
  data visualisation; grand tour; statistical computing; statistical graphics; multivariate data; dynamic graphics
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
  \usepackage{graphicx}
  \usepackage{nicematrix}
  \NiceMatrixOptions{code-for-first-row = \color{red} ,code-for-last-row = \color{red} ,code-for-first-col = \color{blue} ,code-for-last-col = \color{blue}}
output: rticles::tf_article
keep_tex: true
---

@@

<!-- Background -->
From a statistical perspective 3D is a rare data dimension, so unlike in most 3D rotation computer graphics applications, the more useful methods for data analysis need to work for arbitrary dimension. A good approach is to show projections from an arbitrary dimensional space to create dynamic data visualizations called *tours*. Tours involve views of high-dimensional ($p$) data with low-dimensional ($d$) projections. In his original paper on the grand tour, @As85 provided several algorithms for tour paths that could theoretically show the viewer the data *from all sides*. Prior to Asimov's work, there were numerous preparatory developments including @tukey's PRIM-9. PRIM-9 had user-controlled rotations on coordinate axes, allowing one to manually tour through low-dimensional projections. (A video illustrating the capabilities is available through video library of @ASA22.) Steering through all possible projections is impossible, unlike Asimov's tours which allows one to quickly see many, many different projections. After Asimov there have been many tour developments, which are summarized in @lee2021.

@@

Another potentially useful manual control, is to allow the user to choose the position of the center of a slice. The slice tour was introduced in @slicetour. It operates by converting the projection plane into a slice, by removing or de-emphasizing points that are further than a fixed orthogonal distance from the plane. The projection plane is usually thought of as passing through the center of the data. Manual control would allow the user to change the position of the center point, by shifting it along a coordinate axis, while keeping the orientation of the projection plane fixed. The purpose would be to explore how or if the shape of the data, in the space orthogonal to the projection, changes as one gets away from the center. It would also allow the user to interactively decide on the thickness of the slice.

@@

The methods for updating component values in @cook_manual_1997 (and utilized in @spyrison_spinifex_2020) are prescribed primarily for a 2D projection, to take advantage of (then) newly developed 3D trackball controls made available for computer gaming. The first step was to construct a 3D manipulation space from a 2D projection. In this space, the coefficient of the controlled variable ranges between -1 and 1. Movements of a cursor are recorded and converted into changes in the values of $V_m$ thus changing the displayed 2D projection. The algorithm also provided constraints to horizontal, vertical, radial or angular motions only. The construction of the manipulation space overly complicates the manual controls, especially when considering possible techniques that will apply to arbitrary $d$.

@@

The new approach emerged from experiments on the tour using the linear algebra capabilities, and relatively new interactive graphics interface, available in Mathematica [@Mathematica]. The components corresponding to $V_m$ are directly controlled by cursor movement, which updates row $m$ of $A$. The updated matrix is then orthonormalised.

@@

Mathematica provides much utility and versatility, such as inbuilt data visualization, data manipulation and analysis, dynamic functionality, and symbolic and numeric computation. Most of this inbuilt functionality is user-friendly and described in the Mathematica documentation; typically, numerous examples are provided within the documentation, and some possible issues are outlined there.

@@

The examples presented in the application section below (and attached in notebook format in the supplementary material) all use our primary new function, `SliceDynamic`. This function typically accepts grouped data in the form of a matrix where the second last column details the name of the group and the last column details the group index (which can be one if there is only one group). After specifying the initial slice thickness and the slice range, the user is presented with an interactive display in which the control objects appear on the left and the slice visualization appears on the right. The user can change the orientation of the slice via the locator pane, which changes the projection matrix;  a slider controls the slice thickness; and there is an input field, which changes the slice center. The user can also change the appearance of the plot by zooming into the center or changing the point size with the sliders provided. It is worth noting that this zoom works best with data that has been centered and scaled. The projected data can be displayed to contrast it with the slice by ticking a box. The explicit projection matrix can also be displayed via another checkbox. This is especially useful when we wish to import a projection that was identified as interesting in the manual exploration into a later study, as shown in the applications below.

@@

    - `data`: A data matrix
    - `projmat`: The initial projection matrix
    - `centrepoint`: The initial position of the slice
    - `height`: The initial height of the slice
    - `heightRange`: The range of heights
    - `ptSize1`: The size of the points that exist within the slice
    - `ptSize2`: The size of the points that exist outside of the slice

@@

This function uses a different display for the projection that I find very useful. However as it stands it is specific to one data set. Needs to be generalized to arbitrary data if we want to keep it.

@@

We start by exploring the projections of the model prediction. Figure \ref{proj1} summarizes the process. Rotating manually the view, we can visualize the location of predictions for each of the three species, and also get a sense of the difference between the two models. 
To illustrate this difference, we have manually rotated the projection for the RF model (left plot) to identify a view that shows the non-linear but block-type structure that is typical of this type of model. This particular projection ($A_1$) is exported so that it can later be used to show the LDA model (middle plot) and the actual data (right plot). What can be seen is the linearity of the LDA model, where the boundaries are linear and oblique to the variable axes. And, interestingly, this particular projection of the original data shows very distinct clusters of the three species. That means, the obscuring of the boundaries between groups for both of the models is driven by what is happening in the orthogonal space to the plane of the selected projection.

### 2022-10-04 paper/paper.tex
Spellcheck and typos

\begin{abstract}
This paper describes new user controls for examining high-dimensional
data using low-dimensional linear projections and slices. A user can
interactively change the contribution of a given variable to a
low-dimensional projection, which is useful for exploring the
sensitivity of structure to particular variables. The user can also
interactively shift the center of a slice, for example, to explore how
structure changes in local subspaces. Two Mathematica notebooks are
provided, which contains functions enabling the user to experiment with
these new manual controls, with one specifically for exploring regions
and boundaries produced by classification models. The advantage of
Mathematica is its linear algebra capabilities, and interactive cursor
location controls. Some limited implementation has also been made
available in the R package \tt{tourr}. 
\end{abstract}

@@

From a statistical perspective 3D is a rare data dimension, so unlike in
most 3D rotation computer graphics applications, the more useful methods
for data analysis need to work for arbitrary dimension. A good approch
is to show projections from an arbitrary dimensional space to create
dynamic data visualizations called \emph{tours}. Tours involve views of
high-dimensional (\(p\)) data with low-dimensional (\(d\)) projections.
In his original paper on the grand tour, \citet{As85} provided several
algorithms for tour paths that could theoretically show the viewer the
data \emph{from all sides}. Prior to Asimov's work, there were numerous
preparatory developments including \citet{tukey}'s PRIM-9. PRIM-9 had
user-controlled rotations on coordinate axes, allowing one to manually
tour through low-dimensional projections. (A video illustrating the
capabilities is available through video library of \citet{ASA22}.)
Steering through all possible projections is impossible, unlike Asimov's
tours which allows one to quickly see many, many different projections.
After Asimov there have been many tour developments, which are
summarized in \citet{lee2021}.

@@

Another potentially useful manual control, is to allow the user to
choose the position of the center of a slice. The slice tour was
introduced in \citet{slicetour}. It operates by converting the
projection plane into a slice, by removing or de-emphasizing points that
are further than a fixed orthogonal distance from the plane. The
projection plane is usually thought of as passing through the center of
the data. Manual control would allow the user to change the value of
center point, by shifting it along a coordinate axis, while keeping the
orientation of the projection plane fixed. The purpose would be to
explore how or if the shape of the data, in the space orthogonal to the
projection, changes as one gets away from the center. It would also
allow the user to interactively decide on the thickness of the slice.

@@

The methods for updating component values in \citet{cook_manual_1997}
(and utilised in \citet{spyrison_spinifex_2020}) are prescribed
primarily for a 2D projection, to take advantage of (then) newly
developed 3D trackball controls made available for computer gaming. The
first step was to construct a 3D manipulation space from a 2D
projection. In this space, the coefficient of the controlled variable
ranges between -1 and 1. Movements of a cursor are recorded and
converted into changes in the values of \(V_m\) thus changing the
displayed 2D projection. The algorithm also provided constraints to
horizontal, vertical, radial or angular motions only. The construction
of the manipulation space overly complicates the manual controls,
especially when considering possible techniques that will apply to
arbitrary \(d\).

@@

The new approach emerged from experiments on the tour using the linear
algebra capabilities, and relatively new interactive graphics interface,
available in \citet{Mathematica}. The components corresponding to
\(V_m\) are directly controlled by cursor movement, which updates row
\(m\) of \(A\). The updated matrix is then orthonormalised.

@@

The examples presented in the application section below (and attached in
notebook format in the supplementary material) all use our primary new
function, \texttt{SliceDynamic}. This function typically accepts grouped
data in the form of a matrix where the second last column details the
name of the group and the last column details the group index (which can
be one if there is only one group). After specifying the initial slice
thickness and the slice range, the user is presented with an interactive
display in which the control objects appear on the left and the slice
visualisation appears on the right. The user can change the orientation
of the slice via the locator pane, which changes the projection matrix;
a slider controls the slice thickness; and there is an input field,
which changes the slice centre. The user can also change the appearance
of the plot by zooming into the center or changing the point size with
the sliders provided. It is worth noting that this zoom works best with
data that has been centred and scaled. The projected data can be
displayed to contrast it with the slice by ticking a box. The explicit
projection matrix can also be displayed via another checkbox. This is
especially useful when we wish to import a projection that was
identified as interesting in the manual exploration into a later study,
as shown in the applications below.

@@

We start by exploring the projections of the model prediction. Figure
\ref{proj1} summarises the process. Rotating manually the view, we can
visualise the location of predictions for each of the three species, and
also get a sense of the difference between the two models. To illustrate
this difference, we have manually rotated the projection for the RF
model (left plot) to identify a view that shows the non-linear but
block-type structure that is typical of this type of model. This
particular projection (\(A_1\)) is exported so that it can later be used
to show the LDA model (middle plot) and the actual data (right plot).
What can be seen is the linearity of the LDA model, where the boundaries
are linear and oblique to the variable axes. And, interestingly, this
particular projection of the original data shows very distinct clusters
of the three species. That means, the obscuring of the boundaries
between groups for both of the models is driven by what is happening in
the orthogonal space to the plane of the selected projection.

>>>>

\begin{abstract}
This paper describes new user controls for examining high-dimensional
data using low-dimensional linear projections and slices. A user can
interactively change the contribution of a given variable to a
low-dimensional projection, which is useful for exploring the
sensitivity of structure to particular variables. The user can also
interactively shift the center of a slice, for example, to explore how
structure changes in local subspaces. Two Mathematica notebooks are
provided, which contain functions enabling the user to experiment with
these new manual controls, with one specifically for exploring regions
and boundaries produced by classification models. The advantage of
Mathematica is its linear algebra capabilities, and interactive cursor
location controls. Some limited implementation has also been made
available in the R package \tt{tourr}. 
\end{abstract}

@@

From a statistical perspective 3D is a rare data dimension, so unlike in
most 3D rotation computer graphics applications, the more useful methods
for data analysis need to work for arbitrary dimension. A good approach
is to show projections from an arbitrary dimensional space to create
dynamic data visualizations called \emph{tours}. Tours involve views of
high-dimensional (\(p\)) data with low-dimensional (\(d\)) projections.
In his original paper on the grand tour, \citet{As85} provided several
algorithms for tour paths that could theoretically show the viewer the
data \emph{from all sides}. Prior to Asimov's work, there were numerous
preparatory developments including \citet{tukey}'s PRIM-9. PRIM-9 had
user-controlled rotations on coordinate axes, allowing one to manually
tour through low-dimensional projections. (A video illustrating the
capabilities is available through video library of \citet{ASA22}.)
Steering through all possible projections is impossible, unlike Asimov's
tours which allows one to quickly see many, many different projections.
After Asimov there have been many tour developments, which are
summarized in \citet{lee2021}.

@@

Another potentially useful manual control, is to allow the user to
choose the position of the center of a slice. The slice tour was
introduced in \citet{slicetour}. It operates by converting the
projection plane into a slice, by removing or de-emphasizing points that
are further than a fixed orthogonal distance from the plane. The
projection plane is usually thought of as passing through the center of
the data. Manual control would allow the user to change the position of
the center point, by shifting it along a coordinate axis, while keeping
the orientation of the projection plane fixed. The purpose would be to
explore how or if the shape of the data, in the space orthogonal to the
projection, changes as one gets away from the center. It would also
allow the user to interactively decide on the thickness of the slice.

@@

The methods for updating component values in \citet{cook_manual_1997}
(and utilized in \citet{spyrison_spinifex_2020}) are prescribed
primarily for a 2D projection, to take advantage of (then) newly
developed 3D trackball controls made available for computer gaming. The
first step was to construct a 3D manipulation space from a 2D
projection. In this space, the coefficient of the controlled variable
ranges between -1 and 1. Movements of a cursor are recorded and
converted into changes in the values of \(V_m\) thus changing the
displayed 2D projection. The algorithm also provided constraints to
horizontal, vertical, radial or angular motions only. The construction
of the manipulation space overly complicates the manual controls,
especially when considering possible techniques that will apply to
arbitrary \(d\).

@@

The new approach emerged from experiments on the tour using the linear
algebra capabilities, and relatively new interactive graphics interface,
available in Mathematica \citep{Mathematica}. The components
corresponding to \(V_m\) are directly controlled by cursor movement,
which updates row \(m\) of \(A\). The updated matrix is then
orthonormalised.

@@

The examples presented in the application section below (and attached in
notebook format in the supplementary material) all use our primary new
function, \texttt{SliceDynamic}. This function typically accepts grouped
data in the form of a matrix where the second last column details the
name of the group and the last column details the group index (which can
be one if there is only one group). After specifying the initial slice
thickness and the slice range, the user is presented with an interactive
display in which the control objects appear on the left and the slice
visualization appears on the right. The user can change the orientation
of the slice via the locator pane, which changes the projection matrix;
a slider controls the slice thickness; and there is an input field,
which changes the slice center. The user can also change the appearance
of the plot by zooming into the center or changing the point size with
the sliders provided. It is worth noting that this zoom works best with
data that has been centered and scaled. The projected data can be
displayed to contrast it with the slice by ticking a box. The explicit
projection matrix can also be displayed via another checkbox. This is
especially useful when we wish to import a projection that was
identified as interesting in the manual exploration into a later study,
as shown in the applications below.

@@

We start by exploring the projections of the model prediction. Figure
\ref{proj1} summarizes the process. Rotating manually the view, we can
visualize the location of predictions for each of the three species, and
also get a sense of the difference between the two models. To illustrate
this difference, we have manually rotated the projection for the RF
model (left plot) to identify a view that shows the non-linear but
block-type structure that is typical of this type of model. This
particular projection (\(A_1\)) is exported so that it can later be used
to show the LDA model (middle plot) and the actual data (right plot).
What can be seen is the linearity of the LDA model, where the boundaries
are linear and oblique to the variable axes. And, interestingly, this
particular projection of the original data shows very distinct clusters
of the three species. That means, the obscuring of the boundaries
between groups for both of the models is driven by what is happening in
the orthogonal space to the plane of the selected projection.
